Mendelian randomization for causal inference in Genome-wide association studies

### 2022-02-06 mr-review.Rmd
added text around multi-SNP methods

There are several ways that experimental data may violate one or more of the three MR assumptions.
Let's first consider the exclusion restriction assumption. It states that the genetic marker 
must affect the outcome only through the exposure. The causal diagram in Figure \@ref(fig:dag1) illustrates 
a scenario that violates this assumption when the variable X is used as the exposure [@vanderweele_methodological_2014]. Bias in causal effect estimation would result. However, 
if variable W is measured and treated as the exposure, then an unbiased estimate of causal effect is 
possible, since W captures the two pathways, W -> Y and W -> X -> Y, through which the genetic marker G 
affects the outcome Y [@vanderweele_methodological_2014;@textor2011dagitty;@textor2016robust].

@@

Sensitivity analysis is recommended and widely used in MR studies because of the inability to verify 
the three MR assumptions with observational data. The goal of a sensitivity analysis is to gain insight 
into how the results might differ were we able to verify the MR assumptions.

@@

Recognition of complex traits' diverse genetic architectures has fueled development of two-sample MR methods with multiple instruments [@hormozdiari_leveraging_2018;@watanabe_global_2019].

@@

@@

Horizontal pleiotropy refers to the presence of multiple pathways from a genetic 
variant to the outcome. One of these multiple pathways may contain the exposure variable.
The fact that there are multiple pathways from the genetic variant to the outcome violates
the MR assumptions. @verbanck_detection_2018 documented widespread horizontal pleiotropy 
in GWAS. This finding motivated the development of MR-PRESSO and inspired 
further research into methods to accommodate horizontal pleiotropy in MR studies [@verbanck_detection_2018].
* What does MR-PRESSO do? What are its limitations?

>>>>

There are several ways that experimental data may violate one or more of the three MR assumptions.
Let's first consider the exclusion restriction assumption (assumption 2 above). It states that the genetic marker 
must affect the outcome only through the exposure. The causal diagram in Figure \@ref(fig:dag1) illustrates 
a scenario that violates this assumption when the variable X is used as the exposure [@vanderweele_methodological_2014]. Bias in causal effect estimation would result. However, 
if variable W is measured and treated as the exposure, then an unbiased estimate of causal effect is 
possible, since W captures the two pathways, W -> Y and W -> X -> Y, through which the genetic marker G 
affects the outcome Y [@vanderweele_methodological_2014;@textor2011dagitty;@textor2016robust].

@@

Sensitivity analysis is recommended and widely used in MR studies because of the inability to verify 
the three MR assumptions with observational data. The goal of a sensitivity analysis is to gain insight 
into how the results might differ if the assumptions are violated.

@@

Recognition of complex traits' diverse genetic architectures has fueled development of two-sample MR 
methods with multiple SNP instruments [@hormozdiari_leveraging_2018;@watanabe_global_2019].

@@

In the case of continuous exposures and outcomes, we extend the methods and models for 
single-SNP MR to accommodate multiple SNPs. For two stage analyses, we perform 
a sequence of two linear regressions, as in the single-SNP MR models (Equations \@ref(eq:lm3) and \@ref(eq:lm4)). 
For an analysis
with $S$ SNP instruments, one performs two stage analysis by first regressing the 
exposure $n$ vector $X$ on the $n$ by $S$ matrix of SNP instruments. Taking the fitted exposure values 
from this regression, the second regression involves regressing the outcome variable $Y$ on 
the fitted values.

\begin{equation}
  X  = \alpha Z + \epsilon
  (\#eq:lm3)
\end{equation}

where $\epsilon$ is assumed to be a vector of iid normal random variables with mean zero.

\begin{equation}
  Y = \beta \hat X + \tau
  (\#eq:lm4)
\end{equation}

@@

Horizontal pleiotropy refers to the presence of multiple pathways from a genetic 
variant to the outcome. One of these multiple pathways may contain the exposure variable.
The fact that there are multiple pathways from the genetic variant to the outcome violates
the MR assumptions. @verbanck_detection_2018 documented widespread horizontal pleiotropy 
in GWAS. This finding motivated the development of MR-PRESSO and inspired 
further research into methods to model and to account for horizontal pleiotropy [@verbanck_detection_2018].
